---
title: "Introduction to spotr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to spotr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE
   #comment = "#>"
)
```

First load some packages needed for fitting models and producing plots.
```{r setup, message=FALSE}
library(spotr)
library(mgcv)
library(brms)
library(sf)
library(ggplot2)
library(dplyr)
```

Next load the data for the case study. 
They data are distributed with `spotr` and consists of 20 years of monitoring data of the common cuckoo in Sweden.

```{r}
data(cuckoo)
head(cuckoo)
```

Transform the data into an `sp` object for plotting and for spatial manipulation.
Also load the map of Sweden contained in the package.
```{r}
cuckoo = st_as_sf(cuckoo, coords = c("lon", "lat"), crs = st_crs(4326), remove = FALSE)
data(swe_map)
```

Plot the data for an example year
```{r}
ggplot(swe_map) +  geom_sf() + geom_sf(aes(col = count), data = subset(cuckoo, yr == 2010)) + theme_minimal()
```

To compute indices a model that can generate predictions of abundances across time and space is needed.
For example, a tensor product of a temporal and a spatial smooth can be fit with `mgcv`
```{r}
gam_fit = gam(count ~ te(yr, lat, lon, d = c(1,2), k = c(8,30)), data = cuckoo, family = nb())
```

A data frame containing the spatial locations as well as any variables needed for predicting from the model is now needed.
If there are no spatio-temporal explanatory variables, as in the model above, the data frame may be defined from the combination of a data frame containing a spatial grid and a data frame containing temporal variables.

First define a spatial grid. 
```{r}
spat_grid = st_make_grid(cuckoo, cellsize = .5, what = "centers") |> 
  st_intersection(swe_map) |> st_as_sf()
ggplot(swe_map) + geom_sf() + geom_sf(data = spat_grid) + theme_minimal()
```

To avoid predicting far out of the range of the species, one can restrict the spatial grid to the convex hull of the observed range.
The cuckoo occurs throughout Sweden so the effect is small, but it could be important for other species.
```{r}
spat_grid = st_intersection(spat_grid, st_convex_hull(st_union(cuckoo)))
ggplot(swe_map) + geom_sf() + geom_sf(data = spat_grid) + theme_minimal()

# Only need coordinates
spat_grid = as.data.frame(st_coordinates(spat_grid))
colnames(spat_grid) = c("lon", "lat")
```

The grid now contains at least one point for each county in the map, which will be important later when producing county-wise indices.
  
Then define the temporal grid. The model only contains one temporal variable.

```{r}
temp_grid = data.frame(yr = 2000:2020)
```

Now combine the spatial and temporal grids using `merge`.

```{r}
pred_points = merge(temp_grid, spat_grid, by = NULL)
```

The spatio-temporal grid can now be used to compute an index for the total Swedish population.
The result is a data frame with an index for each year and uncertainty estimates.

```{r}
swe_ind = index(gam_fit, timevar = "yr", newdata = pred_points)
head(swe_ind)
```

```{r}
swe_ind |> 
  ggplot(aes(x = yr, y = index, ymin = ci_2.5, ymax = ci_97.5)) + geom_ribbon(fill = "blue", alpha = .3) + geom_line() + theme_minimal()
```

Confidence intervals for the index are fairly wide. 
Part of the reason for this may be that it is hard to estimate population size in the first year when few sites were surveyed.
One can instead use the mean abundance over a sequence of years as the baseline using the `baseline` argument to `index`:

```{r}
index(gam_fit, timevar = "yr", newdata = pred_points, baseline = 2000:2010) |> 
  ggplot(aes(x = yr, y = index, ymin = ci_2.5, ymax = ci_97.5)) + geom_ribbon(fill = "blue", alpha = .3) + geom_line() + theme_minimal()
```

This results in lower uncertainty because the mean over 2000 - 2010 can be more accurately estimated.


